Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-07-04 manuscript/manuscript.Rmd
minor updates.

The now extensive corpus of experimental archaeology have witnessed the growing principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

@@

Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often. One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition. The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity. Roe and Just [@roe2009: 1266-1267] defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings". This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs. In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades [@eren2016]. Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area. All the variables of interest are easy to measure, quantify, and control in a machine knapping setting. Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments. As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].

@@

>>>>

The now extensive corpus of experimental archaeology have witnessed the growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

@@

Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms another major issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades []. Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area. All the variables of interest are easy to measure, quantify, and control in a machine knapping setting. Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in in a series of recently published experimental studies aiming at understanding the transmission of lithic technologies. As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020]. how decontexualized law-like statements interact in a particular context.

@@

Again, actualistic expeirment as a huestric for hypothesis generation as well as a low-cost , echoing the spirit stated by Lin et al. [-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".
